---
title: "Black students are disproprotionately disciplined in Delaware"
---
Get started, loading & cleaning the data.
```{r setup, include=FALSE}
# Turn off scientific notation
options(scipen=999) 
library(tidyverse)
library(janitor)
library(magrittr) # use magrittr if you want to replicate the function of the tidyverse |> character in vscode or a similar program - this code uses the |> character a lot
library(dplyr)
library(ggthemes)
library(stringr)
library(purrr)

# Remove comment from the below read_csv line as necessary.
delaware_discipline <- read_csv("/Users/joellevtov/Library/CloudStorage/GoogleDrive-levtov@terpmail.umd.edu/My Drive/Journalism/delawarediscipline/delaware_student_discipline_raw.csv") |> 
    clean_names() |> 
    select(!geography) #the geography column reads either "All Students" or, for redacted rows, "NA" - so this will ignore the column
delaware_discipline$sub_group <- str_replace(delaware_discipline$race, regex("Asian", ignore_case = TRUE), "Asian American") 
delaware_discipline$sub_group <- str_replace(delaware_discipline$sub_group, regex("Asian", ignore_case = TRUE), "Asian American") 
delaware_discipline$category <- str_replace(delaware_discipline$category, regex("Out-of-School Suspension, No CDAP Placement", ignore_case = TRUE), "Out-of-School Suspension without CDAP Placement")
# The raw data has two different phrases that mean the same thing - "Out-of-School Suspension, No CDAP Placement" and "Out-of-School Suspension without CDAP Placement". For clarity, I replaced the latter with the former.


delaware_pop <- read_csv("/Users/joellevtov/Downloads/Student_Enrollment_20240216.csv") |> 
    clean_names()
delaware_pop$grade <- str_replace(delaware_pop$grade, regex("Twelfth", ignore_case=TRUE), "12th Grade")
delaware_pop$sub_group <- str_replace(delaware_pop$sub_group, regex("Asian", ignore_case = TRUE), "Asian American")
```

Let's make the racial labels easier to type.
```{r}
race_categories <- as.data.frame(unique(na.omit(delaware_discipline$race))) 
colnames(race_categories)[1] <- "race"
race_categories <- as.data.frame(race_categories) |> 
    filter(race != "All Students") #removed the line that had "All Students" as a race
name_hisp <- race_categories[1,1]
name_white <- race_categories[2,1]
name_native_am <- race_categories[3,1]
name_black <- race_categories[4,1]
name_asian_am <- race_categories[5,1]
name_multi_racial <- race_categories[6,1]
name_hawaiian_pi <- race_categories[7,1]
print_var_name_and_value <- function(x) {
  # Capture and print the variable name
  var_name <- deparse(substitute(x))
  
  # Capture and print the variable value
  var_value <- eval(substitute(x))
  
  # Combine into a data frame for pretty printing
  result_df <- data.frame(
    variable_name = var_name,
    variable_value = as.character(var_value),
    stringsAsFactors = FALSE  # Avoid factors for strings
  )
  
  print(result_df)
}

# Run each line individually to make sure that the variables represent the correct string - this will be critical later on.
print_var_name_and_value(name_hisp)
print_var_name_and_value(name_white)
print_var_name_and_value(name_native_am)
print_var_name_and_value(name_black)
print_var_name_and_value(name_asian_am)
print_var_name_and_value(name_multi_racial)
print_var_name_and_value(name_hawaiian_pi)
```

Let's break down the <delaware_discipline> dataframe.
<school_year> shows what school year we are addressing
<district_code> is a column we will ignore, but it shows the number code of each district. The one we want is 0, the state of Delaware.
<district> shows in words the name of each district.
<school_code> shows the number code of each school.
<organization> shows the school name
<race> shows what race the statistics address
<special_demo> is a more clear version of the <sub_group> column. It shows what subgroup the statistics contained in the row are for
<sub_group> shows the above in a more complicated form -- i.e. instead of reading "ESL students", it might read "Hispanic/Latino/1st Grade/Out-of-School Suspension with CDAP/ESL students." This would be helpful, but it doesn't list all the statuses, e.g. what district one is looking at
<category> displays what type of discipline was applied - in-school discipline, out-of-school discipline, expulsion, etc.
<rowstatus> shows whether or not a row has been ommitted out of privacy concerns - if the pool of students is small enough, it would be possible to identify which students were disciplined, in theory
<students> shows the number of STUDENTS who were disciplined with <category>
<enrollment> shows the numer of students who fit into a certain <status>
<pct_enrollment> shows what percent of the student body the studeents in <sub_group> make up
<incidents> shows the number of times <status> was applied - e.g the number of times a student got in-school suspension, NOT the number of students that got in-school suspension
<avg_duration> shows how many days students in <sub_group> were suspended for on average

That dataframe is helpful, but is hard to query for population counts -- taking the median of the number of Hispanic 2nd graders is hard to do because the enrollment numbers are repeated 2-5 times with each discipline <category>. Luckily, the state offers another spreadsheet with exclusively enrollment numbers. 
race                             num_enrolled    num_disciplined    num_incidents   per_capita_disciplines length
Hispanic/Latino                  x               x                  x               x                       x
Native American                  x               x                  x               x                       x
African American                 x               x                  x               x                       x
White                            x               x                  x               x                       x
Asian American                   x               x                  x               x                       x
Multi-Racial                     x               x                  x               x                       x
Native Hawaiian/Pacific Islander x               x                  x               x                       x

<num_enrolled> is the average number of students in a grade across all eight years
<num_disciplined> is the average number of students that have faced discipline in each category
<num_of_incidents> is the average number of incidents (i.e. NOT the number of students)
<per_capita_disciplines> is the average number of students disciplined, i.e. num_students_disciplined/enrollement
<length> is the average length of a discipline measure in days

Now we can start calculating. First, we calculate how many students of each race there are overall -- e.g. on average, how many white 10th graders there were in Delaware. Taking the median here so that we don't overcount the number of students like we'd do if we were to sum or get an average thrown off by outliers like we would get with the median. Big fluctuations in enrollment are certainly an interesting story, but not the one we are focusing on here.
```{r}
# This works well, we just have to keep in mind that Delaware has redacted several rows for small schools, such as the Positive Outcome Charter School - likely for FERPA reasons.
# this function calculates the average number of students of each race, gender, district and sub_group
median_student_num <- function(.data) { 
    filtered_race_data <- .data |> 
        ungroup() |> 
        group_by(race, grade, gender, district_code, sub_group) |> 
        summarize(students = median(na.omit(students)))
    filtered_race_data$school_year <- NA
return(filtered_race_data)
}
filtered_race_data <- median_student_num(delaware_pop)
    
delaware_pop_copy <- delaware_pop |> 
    full_join(filtered_race_data, join_by(grade, sub_group, students, race, district_code, school_year, gender))
```

The process_race_data function is a one-stop shop to calculating everything we could want. One can specify the following parameters:
- District Code (this argument is mandatory)
- Race
- Grade
- Gender
- Discipline type (i.e. in-school suspension, out-of-school suspension, expulsion)

If interested in all students in a particular category, e.g. the average number of students of all races, write "all" (any capitalization) or leave the argument out.
```{r}
process_race_data <- function(pop_df, discipline_df, dis_code, race_name, grade_query, gender_query, disc_type) {
    library(stringr)
    library(tidyverse)

    filter_incidents <- function(.data, dis_code, race_name = NULL, grade_query = NULL, gender_query = NULL, disc_type = NULL) {
    # Start with filtering district_code as it's applied regardless
    filtered <- .data |> 
        filter(district_code == dis_code) |> 
        filter(!category %in% c("Out-of-School Suspension with CDAP Placement", "Out-of-School Suspension without CDAP Placement")) # since delaware does not reliably include the rows about CDAP placement vs. without CDAP placement, I'm going to filter them out
    
    # Apply filter for race_name. Default to "All Students" if "all" or NULL
    if (is.null(race_name) || tolower(race_name) == "all") {
        filtered <- filtered |>  filter(race == "All Students")
    } else {
        filtered <- filtered |> filter(race == race_name) |> filter(sub_group == race_name)
    }
    
    # Apply filter for grade_query. Use str_detect for partial matches. Default to "All Students" if "all" or NULL
    if (is.null(grade_query) || tolower(grade_query) == "all") {
        filtered <- filtered |>  filter(grade == "All Students")
    } else {
        filtered <- filtered |>  filter(str_detect(grade, regex(grade_query, ignore_case = TRUE)))
    }
    
    # Apply filter for gender_query. Default to "All Students" if "all" or NULL
    if (is.null(gender_query) || tolower(gender_query) == "all") {
        filtered <- filtered |>  filter(gender == "All Students")
    } else {
        filtered <- filtered |>  filter(gender == gender_query)
    }

    # Apply filter for disc_type if provided and not "all". No "All Students" default for discipline type as it doesn't apply
    if (!is.null(disc_type) && tolower(disc_type) != "all") {
        filtered <- filtered |>  filter(category == disc_type)
    }

    filtered <- filtered |> 
    summarize(num_incidents=median(na.omit(incidents)), num_disc_students=median(na.omit(students)))

    return(filtered)
    print(filtered)
    }

# this function finds the average number of students of a certain race
find_num_race <- function(.data, dis_code, race_name = NULL, grade_query = NULL, gender_query = NULL, disc_type = NULL) { 
        filtered <- .data |> # .data takes the population dataframe
            filter(district_code== dis_code)
    if (is.null(race_name) || tolower(race_name) == "all") {
        filtered <- filtered |> filter(race == "All Students")
    } else {
        filtered <- filtered |> filter(race == race_name) |> filter(sub_group == race_name)
    }
        if (is.null(grade_query) || tolower(grade_query) == "all") {
        filtered <- filtered |>  filter(grade == "All Students")
    } else {
        filtered <- filtered |>  filter(str_detect(grade, regex(grade_query, ignore_case = TRUE)))
    }
        if (is.null(gender_query) || tolower(gender_query) == "all") {
        filtered <- filtered |>  filter(gender == "All Students")
    } else {
        filtered <- filtered |>  filter(gender == gender_query)
    }

filtered <- filtered |> 
    filter(is.na(school_year)) # |> 
                # select(students)
        return(filtered)
    }


    num_race <- find_num_race(pop_df, dis_code, race_name, grade_query, gender_query)
    # this function calculates the average number of incidents per race
    incidents_per_race <- filter_incidents(discipline_df, dis_code, race_name, grade_query, gender_query, disc_type)
    # this joins the dataframes into one
    incidents_per_race$num_students <- num_race$students
    # this changes the order of the columns
    incidents_per_race <- relocate(incidents_per_race, num_students, .before = 1)
    # this calculates the per capita rate
    incidents_per_race <- incidents_per_race |> 
        mutate(per_capita = as.numeric(num_incidents) / as.numeric(num_students))
}

# The above functions make our life a lot easier - now we just have to plug in the dataframes, the district code and the race.
hisp_incidents <- process_race_data(delaware_pop_copy, delaware_discipline, 0, name_hisp, "all", "all", "all") # could also leave out the "all" arguments all together and provide no arguments for grade_query, gender_query, disc_type as they are automatically set to 'null'
print(hisp_incidents)
native_am_incidents <- process_race_data(delaware_pop_copy, delaware_discipline, 0, name_native_am, "all", "all", "all") 
print(native_am_incidents)
black_incidents <- process_race_data(delaware_pop_copy, delaware_discipline, 0, name_black, "all", "all", "all") 
print(black_incidents)
white_incidents <- process_race_data(delaware_pop_copy, delaware_discipline, 0, name_white, "all", "all", "all") 
print(white_incidents)
asian_am_incidents <- process_race_data(delaware_pop_copy, delaware_discipline, 0, name_asian_am, "all", "all", "all") 
print(asian_am_incidents)
multi_racial_incidents <- process_race_data(delaware_pop_copy, delaware_discipline, 0, name_multi_racial, "all", "all", "all") 
print(multi_racial_incidents)
hawaiian_pi_incidents <- process_race_data(delaware_pop_copy, delaware_discipline, 0, name_hawaiian_pi, "all", "all", "all") 
print(hawaiian_pi_incidents)
```

``` {r}


 incidents_per_race <- filter_incidents(discipline_df, dis_code, race_name, grade_query, gender_query, disc_type)
    # this joins the dataframes into one
    incidents_per_race$num_students <- num_race$students
    # this changes the order of the columns
    incidents_per_race <- relocate(incidents_per_race, num_students, .before = 1)
    # this calculates the per capita rate
    incidents_per_race <- incidents_per_race |> 
        mutate(per_capita = as.numeric(num_incidents) / as.numeric(num_race))


number_of_incidents_hisp <- filter_incidents(delaware_discipline, 0, name_hisp)
print(number_of_incidents_hisp)
```



Let's make a graph.
```{r}
overall <- list()
overall$race <- c(name_hisp, name_native_am, name_black, name_white, name_asian_am, name_multi_racial, name_hawaiian_pi) 

overall <- hisp_incidents |> 
    full_join(native_am_incidents) |> 
    full_join(black_incidents) |> 
    full_join(white_incidents) |> 
    full_join(asian_am_incidents) |> 
    full_join(multi_racial_incidents) |> 
    full_join(hawaiian_pi_incidents) |> 
    mutate(
        race=c(name_hisp, name_native_am, name_black, name_white, name_asian_am, name_multi_racial, name_hawaiian_pi)
    )|> 
    rename(num_enrolled=num_students) |> 
    rename(per_capita_disciplines=per_capita) |> 
    relocate(race, .before = num_enrolled)
print(overall)

overall_graph <- overall |> 
  mutate(race_ordered = fct_reorder(race, per_capita_disciplines)) # Assuming 'race' is the column you want on x-axis

overall_graph <- overall_graph |> 
    ggplot() +
    geom_col(aes(x= race_ordered, y=per_capita_disciplines), fill="darkblue") +
    labs(
        title="Black and multi-racial students are far more likely to face discipline in Delaware.",
        x= "race",
        y="avg"
    ) +
    theme_economist() +
    theme(
    axis.text.x = element_text(angle = 45, hjust=1, vjust=0.9, size=15),  # Adjust text angle, justification, and size
    plot.margin = margin(t = 10, r = 10, b = 40, l = 10, unit = "pt")  # Expand bottom margin
    )
print(overall_graph)
```

What's clear from this data is that Black and multi-racial students are far more likely to be disciplined . But do they get more serious discipline - e.g. out-of-school suspension vs. in-school discipline? Let's dive deeper into the data. First, let's create some variables - there are six types of discipline noted in the data - in-school suspension, out-of-school suspension (with and without CDAP -- "Consortium Discipline Alternative Program" -- placement, used when the student has been punished through school programs without improving and they "routinely and seriously disrupt the classroom"). 
```{r}
name_in_school_sus <- "In-School Suspension" 
name_oos <- "Out-of-School Suspension"
name_oos_wo_cdap <- "Out-of-School Suspension without CDAP Placement"    
name_oos_cdap <- "Out-of-School Suspension with CDAP Placement"   
name_expulsion <-"Expulsion (Permanent Removal from School)"      
```

First, let's tackle in-school suspension. 
```{r}
# create a new function that functions similar to the one above
process_detailed_race_data <- function(pop_df, discipline_df, district_code, race_name, disc_type) {
    library(stringr)
    library(tidyverse)


filter_incidents <- function(.data, dis_code, race_name, disc_type) { 
        filtered1 <- .data |> 
            filter(district_code==dis_code) |> 
            filter(category==disc_type) |> 
            filter(race==race_name) |> 
            filter(sub_group==race_name) |> 
            summarize(num_incidents=median(na.omit(incidents)), num_disc_students=median(na.omit(students)))
    return(filtered1)
    }

# this function finds the average number of students of a certain race
    find_num_race <- function(.data, x, race_name) { 
        filtered <- .data |> # .data takes the population dataframe
            filter(district_code==x) |> 
            filter(race==race_name) |> 
            filter(sub_group==race_name) |> 
            filter(is.na(school_year)) |> 
            select(students)
    return(filtered)
    }

    num_race <- find_num_race(pop_df, district_code, race_name)
    # this function calculates the average number of incidents per race
    incidents_per_race <- filter_incidents(discipline_df, district_code, race_name, disc_type)
    # this joins the dataframes into one
    incidents_per_race$num_students <- num_race$students
    # this changes the order of the columns
    incidents_per_race <- relocate(incidents_per_race, num_students, .before = 1)
    # this calculates the per capita rate
    incidents_per_race <- incidents_per_race |> 
        mutate(per_capita = as.numeric(num_incidents) / as.numeric(num_race))

return(incidents_per_race)
}

hisp_in_school_sus <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_hisp, name_in_school_sus)
print(hisp_in_school_sus)
native_am_in_school_sus <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_native_am, name_in_school_sus)
print(native_am_in_school_sus)
black_in_school_sus <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_black, name_in_school_sus)
print(black_in_school_sus)
white_in_school_sus <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_white, name_in_school_sus)
print(white_in_school_sus)
asian_am_in_school_sus <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_asian_am, name_in_school_sus)
print(asian_am_in_school_sus)
multi_racial_in_school_sus <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_multi_racial, name_in_school_sus)
print(multi_racial_in_school_sus)
hawaiian_pi_in_school_sus <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_hawaiian_pi, name_in_school_sus)
print(hawaiian_pi_in_school_sus)

in_school_suspensions <- hisp_in_school_sus |> 
    full_join(native_am_in_school_sus) |> 
    full_join(black_in_school_sus) |> 
    full_join(white_in_school_sus) |> 
    full_join(asian_am_in_school_sus) |> 
    full_join(multi_racial_in_school_sus) |> 
    full_join(hawaiian_pi_in_school_sus)
in_school_suspensions$race <- c(name_hisp, name_native_am, name_black, name_white, name_asian_am, name_multi_racial, name_hawaiian_pi) 
in_school_suspensions <- in_school_suspensions |>  relocate(race, .before=num_students)
print(in_school_suspensions)


in_school_suspensions_graph <- in_school_suspensions |> 
  mutate(race_ordered = fct_reorder(race, per_capita)) # Assuming 'race' is the column you want on x-axis

in_school_suspensions_graph <- in_school_suspensions_graph |> 
    ggplot() +
    geom_col(aes(x= race_ordered, y=per_capita), fill="darkblue") +
    labs(
        title="Black and multi-racial students are far more likely to face in-school discipline in Delaware.",
        x= "race",
        y="avg"
    ) +
    theme_economist() +
    theme(
    axis.text.x = element_text(angle = 45, hjust=1, vjust=0.9, size=15),  # Adjust text angle, justification, and size
    plot.margin = margin(t = 10, r = 10, b = 40, l = 10, unit = "pt")  # Expand bottom margin
    )
print(in_school_suspensions_graph)
```

Very interesting. Now, let's see about out-of-school suspensions with CDAP.
```{r}
hisp_oos_cdap <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_hisp, name_oos_cdap)
print(hisp_oos_cdap)
native_am_oos_cdap <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_native_am, name_oos_cdap)
print(native_am_oos_cdap)
black_oos_cdap <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_black, name_oos_cdap)
print(black_oos_cdap)
white_oos_cdap <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_white, name_oos_cdap)
print(white_oos_cdap)
asian_am_oos_cdap <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_asian_am, name_oos_cdap) # this returns na, as there are no rows which document out-of-school suspensions - with or without CDAP for Asian American students. Presumably, that means there were no out-of-school suspensions for all 6,111 Asian students on average. 
print(asian_am_incidents)
multi_racial_oos_cdap <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_multi_racial, name_oos_cdap)
hawaiian_pi_oos_cdap <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_hawaiian_pi, name_oos_cdap)

out_of_school_cdap_suspensions <- hisp_oos_cdap |> 
    full_join(native_am_oos_cdap) |> 
    full_join(black_oos_cdap) |> 
    full_join(white_oos_cdap) |> 
    full_join(asian_am_oos_cdap) |> 
    full_join(multi_racial_oos_cdap) |> 
    full_join(hawaiian_pi_oos_cdap)
out_of_school_cdap_suspensions$race <- c(name_hisp, name_native_am, name_black, name_white, name_asian_am, name_multi_racial, name_hawaiian_pi) 
out_of_school_cdap_suspensions <- out_of_school_cdap_suspensions |>  relocate(race, .before=num_students)
print(out_of_school_cdap_suspensions)

# this will transform the row with only 'NA' values, asian_am, to 0 so r can sort it without throwing an error.
out_of_school_cdap_suspensions_graph <- out_of_school_cdap_suspensions |> 
  mutate(
      num_incidents = if_else(is.na(num_incidents), 0, num_incidents),
      num_disc_students = if_else(is.na(num_disc_students), 0, num_disc_students),
      per_capita = if_else(is.na(per_capita), 0, per_capita)
    )
  out_of_school_cdap_suspensions_graph <- out_of_school_cdap_suspensions_graph %>%
    mutate(race_ordered = fct_reorder(race, per_capita))


out_of_school_cdap_suspensions_graph <- out_of_school_cdap_suspensions_graph |> 
    ggplot() +
    geom_col(aes(x= race_ordered, y=per_capita), fill="darkblue") +
    labs(
        title="Native American and Native Hawaiian students are most likely to face out-of-school discipline without CDAP in Delaware.",
        x= "race",
        y="avg"
    ) +
    theme_economist() +
    theme(
    axis.text.x = element_text(angle = 45, hjust=1, vjust=0.9, size=15),  # Adjust text angle, justification, and size
    plot.margin = margin(t = 10, r = 10, b = 40, l = 10, unit = "pt")  # Expand bottom margin
    )
print(out_of_school_cdap_suspensions_graph)
```
Let's see about out-of-school suspensions without CDAP.
```{r}
hisp_oos_wo_cdap <- process_detailed_race_data(delaware_pop_copy, delaware_discipline, 0, name_hisp, name_oos_wo_cdap)
native_am_oos_wo_cdap <- statewide_only_discipline |> 
    filter(race==name_native_am) |> 
    filter(category==name_oos_wo_cdap) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
black_oos_wo_cdap <- statewide_only_discipline |> 
    filter(race==name_black) |> 
    filter(category==name_oos_wo_cdap) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
white_oos_wo_cdap <- statewide_only_discipline |> 
    filter(race==name_white) |> 
    filter(category==name_oos_wo_cdap) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
asian_am_oos_wo_cdap <- statewide_only_discipline |> 
    filter(race==name_asian_am) |> 
    filter(category==name_oos_wo_cdap) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
multi_racial_oos_wo_cdap <- statewide_only_discipline |> 
    filter(race==name_multi_racial) |> 
    filter(category==name_oos_wo_cdap) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
hawaiian_pi_oos_wo_cdap <-statewide_only_discipline |> 
    filter(race==name_hawaiian_pi) |> 
    filter(category==name_oos_wo_cdap) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
oos_wo_cdap$num_of_incidents <- c(hisp_oos_wo_cdap, native_am_oos_wo_cdap, black_oos_wo_cdap, white_oos_wo_cdap, asian_am_oos_wo_cdap, multi_racial_oos_wo_cdap, hawaiian_pi_oos_wo_cdap)
oos_wo_cdap <- oos_wo_cdap |> 
    mutate(per_capita = (as.numeric(num_of_incidents)/as.numeric(num_enrolled))*100)
oos_wo_cdap_graph <- oos_wo_cdap
oos_wo_cdap_graph <- oos_wo_cdap_graph |> 
    arrange(desc(per_capita))
oos_wo_cdap_graph |> 
    ggplot() +
    geom_col(aes(x=reorder(race, per_capita), y=per_capita), fill="darkblue") +
    labs(
        title="Native - not Black - students far more likely to face out-of-school suspension without CDAP placement",
        x="race",
        y="avg"
    ) +
    theme_economist()
```

```{r}
expulsion <- demographics_detailed
expulsion$method <- name_expulsion
hisp_expulsion <- statewide_only_discipline |> 
    filter(race==name_hisp) |> 
    filter(category==name_expulsion) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
native_am_expulsion <- statewide_only_discipline |> 
    filter(race==name_native_am) |> 
    filter(category==name_expulsion) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
black_expulsion <- statewide_only_discipline |> 
    filter(race==name_black) |> 
    filter(category==name_expulsion) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
white_expulsion <- statewide_only_discipline |> 
    filter(race==name_white) |> 
    filter(category==name_expulsion) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
asian_am_expulsion <- statewide_only_discipline |> 
    filter(race==name_asian_am) |> 
    filter(category==name_expulsion) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
multi_racial_expulsion <- statewide_only_discipline |> 
    filter(race==name_multi_racial) |> 
    filter(category==name_expulsion) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
hawaiian_pi_expulsion <- statewide_only_discipline |> 
    filter(race==name_hawaiian_pi) |> 
    filter(category==name_expulsion) |> 
    filter(rowstatus=="REPORTED") |> 
    summarize(count=n())
expulsion$num_of_incidents <- c(hisp_expulsion, native_am_expulsion, black_expulsion, white_expulsion, asian_am_expulsion, multi_racial_expulsion, hawaiian_pi_expulsion)
expulsion <- expulsion |> 
    mutate(per_capita = (as.numeric(num_of_incidents)/as.numeric(num_enrolled))*100)
expulsion_graph <- expulsion
expulsion_graph <- expulsion_graph |> 
    arrange(desc(per_capita))
expulsion_graph |> 
    ggplot() +
    geom_col(aes(x=reorder(race, per_capita), y=per_capita), fill="darkblue") +
    labs(
        title="Black students far more likely to face expulsion",
        x="race",
        y="avg"
    ) +
    theme_economist()

```

race                num_enrolled    num_disciplined     per_capita
hispanic            x               x                   x
native american     x               x                   x
african american    x               x                   x
white               x               x                   x
multi-racial        x               x                   x
hawaiian/pi         x               x                   x
